---
title: "Airline Delays"
subtitle: "Assignment 5A: Approach"
author: "Jocelyn Slater"
format: html
editor: visual
---

```{r load-packages, message=FALSE, echo=FALSE}
library(tidyverse)
library(readr)
library(data.table)
library(ggplot2)
```

## Task

Submitted data file with at least some count analysis (+45)

Recreated File in Same Format as given, including missing data where there were empty cells in source data. (+5)

Placed file in Internet-accessible location, such as a publicly-accessible GitHub repo.  Also OK (if less scaleable) to generated and populated dataframe from code. (+5)

Provided code to populate missing data. (+5)

Transformed data from wide to long format. (+10)

Compared percentage (not just counts) of either delays or arrival rates for two airlines overall. \[Either with a chart, or a table; both would be exemplary\].  Include text summarizes findings from this comparison. (+10)

Compared percentage (not just counts) of either delays or arrival rates for two airlines across five cities. \[Either with a chart, or a table; both would be exemplary\].  Include text summarizes findings from this comparison. (+10)

Describe discrepancy between comparing two airlines’ flight performances city-by-city and overall. (+5)

Explain discrepancy between comparing two airlines’ flight performances city-by-city and overall.(+5)

```{r load-in-data}
url <- "https://raw.githubusercontent.com/jocslater-code/DATA607/refs/heads/main/Assignment5A/AirlineDelays.csv"
df_raw <- read.csv(url)
df_raw
```

The chart above describes arrival delays for two airlines across five destinations. Your task is to:
(1) Create a .CSV file (or optionally, a MySQL database!) that includes all of the information above.
You’re encouraged to use a “wide” structure similar to how the information appears above, so
that you can practice tidying and transformations as described below.
(2) Read the information from your .CSV file into R, and use tidyr and dplyr as needed to tidy
and transform your data.
(3) Perform analysis to compare the arrival delays for the two airlines.

## Outline

- Load in the raw airline data
- Tidy the raw data
  - transform from wide to long
  - Fill in missing data
  - Fix up column names
- Calculate percentage of delays for each airlline and location
  - Show chart - great table
  - Show plot 
- Analysis
  - Describe and explain discrepancy between airlines city by city and overall

## Anticipated Challenges
- Reformatting wide to long in R is a new process for me

## Submission Requirements

-   Post all code (SQL, R, Quarto) in a GitHub repository

-   Submit the GitHub link

-   RPubs publication
