---
title: "Week 3B Assignment: Code Base"
subtitle: "Window Functions"
author: "Jocelyn Slater"
format: html
editor: visual
---

```{r load-packages, message=FALSE, echo=FALSE}
library(tidyverse)
library(readr)
library(data.table)
library(ggplot2)
library(janitor)
library(lubridate)
library(zoo)
```

## Introduction

WRITE INTRO

Task:

1.  Find (or ask an LLM to generate!) a dataset that includes time series for two or more separate items.  For example, you could use end of day stock or cryptocurrency prices since Jan 1, 2022 for several instruments.

2.  Use window functions (in SQL or dplyr) to calculate the year-to-date average and the six-day moving averages for each item.

## Data Import

Data is imported from the NOAA website [Climate Data Online (CDO)](https://www.ncei.noaa.gov/cdo-web/). I filtered the data to just include a 4 month record of average daily maximum and minimum temperatures in Central Park through the website before exporting. Because of this since there is not much extraneous data to subset out, there is minimal tidying to do, but for good measure, I ran the janitor packaged and cleaned up the location columns in the data frame. 

```{r import-data, echo=FALSE}
# Import and tidy
url <- "https://raw.githubusercontent.com/jocslater-code/DATA607/refs/heads/main/Week3B/NOAA_temp_data.csv"
raw_noaa_df <- read_csv(url, show_col_types = FALSE)
raw_noaa_df<- clean_names(raw_noaa_df)

noaa_df<- select(raw_noaa_df, date, dly_tmax_normal, dly_tmin_normal)
noaa_df <- noaa_df %>% rename(temp_max = dly_tmax_normal)
noaa_df <- noaa_df %>% rename(temp_min = dly_tmin_normal)

```

```{r}
# Reformat date column for easy handling and make sure things are in proper order
noaa_df <- noaa_df %>%
  mutate(date = dmy(paste0(date, "-2010"))) %>%
  arrange(date)
```

## Window Functions

```{r}
noaa_df <- noaa_df %>%
  # Calculate cumulative averages
  mutate(
    ytd_avg_min = cummean(temp_min),
    ytd_avg_max = cummean(temp_max)
  ) %>%
  # Calculate 6-day moving averages
  mutate(
    mavg_6day_min = rollmean(temp_min, k = 6, fill = NA, align = "right"),
    mavg_6day_max = rollmean(temp_max, k = 6, fill = NA, align = "right")
  )

```

## Results
NOTE - 6 day rolling average means there are leading NaNs because it takes six days to get six days

```{r plot-mavg}
ggplot(noaa_df, aes(x = date)) +
  # Unprocessed Data
  geom_line(aes(y = temp_max, color = "Max Temp (Unprocessed)"), linetype = "dotted", linewidth = 0.8) +
  geom_line(aes(y = temp_min, color = "Min Temp (Unprocessed)"), linetype = "dotted", linewidth = 0.8) +
  
  # 6 Day Moving Average
  geom_line(aes(y = mavg_6day_max, color = "Max Temp (6-day MAvg)"), linetype = "solid", linewidth = 1.1) +
  geom_line(aes(y = mavg_6day_min, color = "Min Temp (6-day MAvg)"), linetype = "solid", linewidth = 1.1) +
  
  # Colors
  scale_color_manual(
    values = c(
      "Max Temp (Unprocessed)"      = "firebrick1",
      "Max Temp (6-day MAvg)" = "firebrick4",
      "Min Temp (Unprocessed)"      = "lightblue3",
      "Min Temp (6-day MAvg)" = "midnightblue"
    ),
    breaks = c("Max Temp (Unproce.)", "Max Temp (6-day MAvg)", "Min Temp (Unproc.)", "Min Temp (6-day MAvg)")
  ) +
  
  # Labeling
  labs(
    title = "2010 Daily Temperatures",
    subtitle = "Dotted = Unprocessed | Solid = 6-Day Moving Average",
    x = "Date",
    y = "Temperature (F)",
    color = "Series"
  ) +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  theme_minimal() +
  theme(
    legend.position = "bottom"
  )

```

```{r}
ggplot(noaa_df, aes(x = date)) +
  geom_line(aes(y = ytd_avg_max, color = "Max Temp (YTD Avg)"), linetype = "solid", linewidth = 0.9) +
  geom_line(aes(y = ytd_avg_min, color = "Min Temp (YTD Avg)"), linetype = "solid", linewidth = 0.9) +
  
  # Colors
  scale_color_manual(
    values = c(
      "Max Temp (YTD Avg)"  = "coral2",
      "Min Temp (YTD Avg)"  = "steelblue2"
    ),
    breaks = c(
      "Max Temp (YTD Avg)", "Min Temp (YTD Avg)")
  ) +
  
  # Labels
  labs(
    title = "2010 YTD Average Temperatures",
    x = "Date",
    y = "Temperature (F)",
    color = "Series"
  ) +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    panel.grid.minor = element_blank()
  ) +
  guides(color = guide_legend(ncol = 2))

```


## Summary and Next Steps

I quickly found that the IMF data was not regular enough for my application and pivoted to weather data from NOAA. In required me to "order" some free data through their online portal which was very easy. Did filtering online
https://www.ncei.noaa.gov/cdo-web/
https://www.ncei.noaa.gov/cdo-web/datasets/NORMAL_DLY/stations/GHCND:USW00094728/detail

Consistent gap between the daily high and lows

## Submission Requirements

-   Post all code (SQL, R, Quarto) in a GitHub repository

    -   Make sure data is hosted on git or through IMF website

-   Submit the GitHub link

-   RPubs publication
