---
title: "Week 1 Assignment: Approach"
author: "Jocelyn Slater"
format: html
editor: visual
---

```{r load-packages, message=FALSE, echo=FALSE}
library(tidyverse)
library(readr)
```

## Introduction

I will be using [this](https://data.cityofnewyork.us/Environment/2015-Street-Tree-Census-Tree-Data/uvpi-gqnh/about_data) tree data set from the City of New York for my first assignment to explore the relationship between borough and tree health. As a New Yorker myself, I am curious to see how location, tree density, and tree health are correlated.

```{r}
# This will read in the first 1000 lines of the data set. 1000 lines is a limit set by the city website
url <- "https://data.cityofnewyork.us/resource/uvpi-gqnh.csv"
data_raw <- read_csv(url)
head(data_raw)
```

My approach to tidying up the data will be to:

-   Subselect columns that strictly relate to borough and tree health

-   Create new summary column combining different tree health factors. New column with have a boolean value to indicate healthy or unhealthy trees

-   Eliminate rows that are missing key information

I anticipate there being some challenge determining which columns are pertinent to the tree health and which are not.

## Conclusion

-   what to do to extend/verify/update my work
-   read in all of the data
