---
title: "Week 3A Assignment"
subtitle: "Global Baseline Estimate"
author: "Jocelyn Slater"
format: revealjs
editor: visual
---

```{r load-packages, message=FALSE, echo=FALSE}
library(tidyverse)
library(readr)
library(data.table)
library(ggplot2)
library(dplyr)
library(gt)
library(gtExtras)
library(purrr)
library(tidyr)
```

## Introduction
- Utilizing movie survey data from week 2
- Global baseline estimate (GBE) system as below: 

::: {style="font-size: 80%;"}
$$\hat{r}_{ui} = \mu + b_u + b_i$$
where 

$\hat{r}_{ui}$ = predicted rating for movie $i$ and user $u$

$\mu$ = mean rating for movie $i$ across all users

$b_u$ = user bias ($\mu_u - \mu$ how this user rated movies compared to the average)

$b_i$ = movie bias ($\mu_i - \mu$ how this movie is rated compared to the average)
:::

## Data Import and Tidying

::: {style="font-size: 80%;"}
-   Data is imported from SQL database

```{r}
#| echo: false
#| eval: true
rating_raw <- read_csv("https://raw.githubusercontent.com/jocslater-code/DATA607/refs/heads/main/Week2/movie_rating_data.csv", show_col_types = FALSE)
glimpse(rating_raw)
```

-   Also make into dataframe that includes NaN ratings for unseen movies

```{r}
#| echo: true
#| eval: true

full_ratings_df <- rating_raw %>%
  complete(user_name, movie_title, fill = list(rating = NaN))
glimpse(full_ratings_df)
```
:::

## Table Formatting
```{r, echo=TRUE}
# Reshape dataframe and print with NaN cells highlighted
#| label: tbl-ratings-gt
#| tbl-cap: "User Ratings Matrix"

# Reshape data into 5 rows (users) and 6 movie columns
ratings_matrix <- full_ratings_df |>
  pivot_wider(
    names_from = movie_title,
    values_from = rating
  )

# 2. Initialize base table
tbl <- ratings_matrix |>
  gt(rowname_col = "user_name") |>
  sub_missing(missing_text = "NaN") |>
  tab_options(
    table.font.size = px(26),  
    table.width = pct(100),             
    data_row.padding = px(12)        
  )

# 3. Apply background fill to missing cells per column
movie_cols <- setdiff(names(ratings_matrix), "user_name")

for (col in movie_cols) {
  # Get row indices where the current column has NA/NaN
  na_rows <- which(is.na(ratings_matrix[[col]]))
  
  if (length(na_rows) > 0) {
    tbl <- tbl |>
      tab_style(
        style = cell_fill(color = "#f8d7da"), 
        locations = cells_body(
          columns = all_of(col),
          rows = na_rows
        )
      )
  }
}
```

## Conditionally Highlighted Table
```{r}
tbl
```

## Global Baseline Estimate Calculation

```{r, echo=TRUE }
# Convert to matrix 
mat <- as.matrix(ratings_matrix[, -1])
rownames(mat) <- ratings_matrix$user_name

# Compute necessary GBE metrics
mu <- mean(mat, na.rm = TRUE)                            # Overall mean movie rating
movie_bias <- colMeans(mat, na.rm = TRUE) - mu           # Movie relative rating (b_i)
user_bias  <- rowMeans(mat, na.rm = TRUE) - mu           # User relative rating (b_u)

# Create imputed matrix
imputed_mat <- mat

# If the cell is NaN (user has not seen the movie), perform GBE calculation and fill in the value
for (user in rownames(mat)) {
  for (movie in colnames(mat)) {
    if (is.na(mat[user, movie])) {
      est <- mu + movie_bias[movie] + user_bias[user]
      imputed_mat[user, movie] <- round(est, 2)
    }
  }
}

# Convert back to data frame, add back in user names
imputed_ratings_df <- as.data.frame(imputed_mat) |>
  tibble::rownames_to_column(var = "user_name")
```

## Global Baseline Estimate Results
- Recommendation threshold (4.5?)
```{r}
# Display table with imputed values

# 1. Start with the imputed data frame
tbl_imputed <- imputed_ratings_df |>
  gt(rowname_col = "user_name") |>
    tab_options(
      table.font.size = px(26),  
      table.width = pct(100),             
      data_row.padding = px(12)        
  )

# 2. Extract column names (excluding user_name)
movie_cols <- setdiff(names(imputed_ratings_df), "user_name")

# 3. Loop over each column and apply background fill ONLY to originally missing rows
for (col in movie_cols) {
  # Identify row indices that were NA/NaN in the original matrix
  na_rows <- which(is.na(ratings_matrix[[col]]))
  
  if (length(na_rows) > 0) {
    tbl_imputed <- tbl_imputed |>
      tab_style(
        style = cell_fill(color = "#e0f2fe"), # Soft blue highlight for imputed values
        locations = cells_body(
          columns = all_of(col),
          rows = na_rows
        )
      )
  }
}

tbl_imputed
```

## Next Steps
- Expand the survey
- Create interface that automatically presents recommendations when the survey is completed
- Experiment with recommendation threshold




