---
title: "Week 3A Assignment: Code Base"
author: "Jocelyn Slater"
format: html
editor: visual
---

```{r load-packages, message=FALSE, echo=FALSE}
library(tidyverse)
library(readr)
library(data.table)
library(ggplot2)
library(dplyr)
library(gt)
library(gtExtras)
library(purrr)
library(tidyr)
```

## Introduction

WRITE INTRO

Task:

Using the information you collected on movie ratings, implement a Global Baseline Estimate recommendation system in R.   The attached spreadsheet provides the implementation algorithm.

Most recommender systems use personalized algorithms like “content management” and “item-item collaborative filtering.” Sometimes non-personalized recommenders are also useful or necessary. One of the best non-personalized recommender system algorithms is the “Global Baseline Estimate.

The job here is to use the survey data collected and write the R code that makes a movie recommendation using the Global Baseline Estimate algorithm.  Please see the attached spreadsheet for implementation details.

## Data Import

Data is imported from SQL database

```{r}
rating_raw <- read_csv("https://raw.githubusercontent.com/jocslater-code/DATA607/refs/heads/main/Week2/movie_rating_data.csv", show_col_types = FALSE)
glimpse(rating_raw)
```

## Data Tidying

```{r}
# Make into datafrane that includes NaN ratings for unseen movies
full_ratings_df <- rating_raw %>%
  complete(user_name, movie_title, fill = list(rating = NaN))
```


```{r}
# Reshape dataframe and print with NaN cells highlighted
#| label: tbl-ratings-gt
#| tbl-cap: "User Ratings Matrix"

# Reshape data into 5 rows (users) and 6 movie columns
ratings_matrix <- full_ratings_df |>
  pivot_wider(
    names_from = movie_title,
    values_from = rating
  )

# 2. Initialize base table
tbl <- ratings_matrix |>
  gt(rowname_col = "user_name") |>
  sub_missing(missing_text = "NaN") #

# 3. Apply background fill to missing cells per column
movie_cols <- setdiff(names(ratings_matrix), "user_name")

for (col in movie_cols) {
  # Get row indices where the current column has NA/NaN
  na_rows <- which(is.na(ratings_matrix[[col]]))
  
  if (length(na_rows) > 0) {
    tbl <- tbl |>
      tab_style(
        style = cell_fill(color = "#f8d7da"), # Soft red fill
        locations = cells_body(
          columns = all_of(col),
          rows = na_rows
        )
      )
  }
}

tbl
```

## Global Baseline Estimate

-   Calculate metrics necessary for global baseline estimate
    -   user avg
    -   mean movie
    -   movie average
-   Calculate estimate (Global Baseline Estimate = Mean Movie Rating + this movie's rating relative to average + this user's rating relative to average)
    -   print table again, with new estimated values filled in (maybe a different color for emphasis?)

Summary statistics for each movie are calculated below:

```{r }
  
# Convert to matrix for calculation convenience (excluding user_name column)
mat <- as.matrix(ratings_matrix[, -1])
rownames(mat) <- ratings_matrix$user_name

# 2. Compute Global Baseline components
mu <- mean(mat, na.rm = TRUE)                            # Overall mean rating
movie_bias <- colMeans(mat, na.rm = TRUE) - mu           # Movie relative rating (b_i)
user_bias  <- rowMeans(mat, na.rm = TRUE) - mu           # User relative rating (b_u)

# 3. Create imputed matrix
imputed_mat <- mat

for (user in rownames(mat)) {
  for (movie in colnames(mat)) {
    if (is.na(mat[user, movie])) {
      # Baseline formula: Overall Mean + Movie Bias + User Bias
      est <- mu + movie_bias[movie] + user_bias[user]
      imputed_mat[user, movie] <- round(est, 2)
    }
  }
}

# 4. Convert back to data frame
imputed_ratings_df <- as.data.frame(imputed_mat) |>
  tibble::rownames_to_column(var = "user_name")
```

```{r}
# Display table with imputed values

# 1. Start with the imputed data frame
tbl_imputed <- imputed_ratings_df |>
  gt(rowname_col = "user_name")

# 2. Extract column names (excluding user_name)
movie_cols <- setdiff(names(imputed_ratings_df), "user_name")

# 3. Loop over each column and apply background fill ONLY to originally missing rows
for (col in movie_cols) {
  # Identify row indices that were NA/NaN in the original matrix
  na_rows <- which(is.na(ratings_matrix[[col]]))
  
  if (length(na_rows) > 0) {
    tbl_imputed <- tbl_imputed |>
      tab_style(
        style = cell_fill(color = "#e0f2fe"), # Soft blue highlight for imputed values
        locations = cells_body(
          columns = all_of(col),
          rows = na_rows
        )
      )
  }
}

tbl_imputed
```
In order to make a recommendation, I need to decide which estimated rating will be the cutoff. I have decided 4.5 to be that cutoff. For any movie that has a global baseline estimate of below 4.5, I will not recommend, but at or above 4.5 I will recommend for that user.

Therefore, I would recommend all of the listed movies to every user in my system, with the exception of Cody Slater and Spiderman Brand New Day. 

## Summary and Next Steps
- Deploy this on a much larger data set
- Build some type of interface to allow a user to create a profile, enter in their ratings, and get recommendations along the lines of "if you liked this, you might also like this"
- Because I used popular movies that have been widely liked, the ratings skewed high. I would like to include more mediocre/unpopular movies. 
- Play around with ratings cutoff for a recommendation


## LLM Model Citations

Google DeepMind. (2026). Gemini 3.6 Flash \[Large language model\]. https://gemini.google.com. Accessed Sept 16, 2026. Transcript available in Github as 3AGeminiTranscript.pdf

