---
title: "Week 2A Assignment: Approach"
author: "Jocelyn Slater"
format: html
editor: visual
---

```{r load-packages, message=FALSE, echo=FALSE}
library(tidyverse)
library(readr)
library(data.table)
library(ggplot2)
library(dplyr)
```

## Introduction
Collect simple movie-rating data, store it in a SQL database, and analyze it in R.

## SQL Code
This SQL code is documented here for completeness. R is not directly interfacing with SQL. The SQL code was originally run in pgAdmin 4 to create the database that was then exported as a .csv.

```sql
-- Users Table
CREATE TABLE users (
    user_id SERIAL PRIMARY KEY,
    name VARCHAR(100) NOT NULL
);

-- Movies Table
CREATE TABLE movies (
    movie_id SERIAL PRIMARY KEY,
    title VARCHAR(255) NOT NULL
);

-- Rating Table 
CREATE TABLE ratings (
    user_id INT REFERENCES users(user_id) ON DELETE CASCADE,
    movie_id INT REFERENCES movies(movie_id) ON DELETE CASCADE,
    rating NUMERIC(3, 1) NOT NULL,
    PRIMARY KEY (user_id, movie_id),
    CONSTRAINT chk_rating CHECK (rating BETWEEN 0.0 AND 5.0)
);

--Insert Users
INSERT INTO users (name) VALUES 
('Heidi Smith'),
('Michael Pento'),
('Cody Slater'),
('Jocelyn Slater')
('Jacob OBrien');

-- Insert Movies
INSERT INTO movies (title) VALUES 
('Spiderman Brand New Day'),
('The Odyssey'),
('Toy Story 5'),
('Project Hail Mary'),
('Superman'), 
('Goat');

-- Insert Ratings
INSERT INTO ratings (user_id, movie_id, rating) VALUES 
(1, 3, 4.3), -- Heidi rated Toy Story 5 4.3
(1, 6, 4.5), -- Heidi rated Goat 4.5
(2, 3, 4.3), -- Michael rated Toy Story 5 a 4.3
(2, 6, 4.0), -- Michael rated Goat 4.0
(2, 4, 4.8), -- Michael rated Project Hail Mary 4.8
(3, 2, 4.5), -- Cody rated The Odyssey 4.5
(3, 3, 3.9), -- Cody rated Toy Story 5 3.9
(3, 4, 3.7), -- Cody rated Project Hail Mary 3.7
(3, 5, 4.2), -- Cody rated Superman 4.2
(3, 6, 3.5), -- Cody rated Goat 3.5
(4, 3, 4.2), -- Jocelyn rated Toy Story 5 4.2
(4, 4, 4.8), -- Jocelyn rated Project Hail Mary 4.8
(4, 5, 4.7), -- Jocelyn rated Superman 4.7
(4, 6, 4.0), -- Jocelyn rated Goat 4.0
(5, 1, 4.6), -- Jacob rated Spideman 4.6
(5, 2, 5.0), -- Jacob rated The Odyssey 5.0
(5, 3, 4.0), -- Jacob rated Toy Story 5 4.0
(5, 4, 4.3), -- Jacob rated Project Hail Mary 4.3
(5, 5, 4.5), -- Jacob rated Superman 4.5
(5, 6, 3.3); -- Jacob rated Goat 3.3
```


## Data Import
Data is imported from SQL database
```{r}
rating_raw <- read_csv("https://raw.githubusercontent.com/jocslater-code/DATA607/refs/heads/main/Week2/movie_rating_data.csv", show_col_types = FALSE)
glimpse(rating_raw)
```
## Data Tidying
```{r}
# Make into datafrane that includes NaN ratings for unseen movies
full_ratings_df <- rating_raw %>%
  complete(user_name, movie_title, fill = list(rating = NaN))
```
   
## Initial Data Analysis   

Summary statistics for each movie are calculated below:
```{r}
movie_summary <- full_ratings_df %>%
  filter(!is.na(rating)) %>% # Exclude NaN / missing ratings
  group_by(movie_title) %>%
  summarise(
    ratings_count = n(),
    avg_rating = mean(rating),
    sd_rating = sd(rating, na.rm = TRUE) 
  )

print(movie_summary)
```

Summary statistics for each user are calculated below:
```{r}
user_summary <- full_ratings_df %>%
  filter(!is.na(rating)) %>%
  group_by(user_name) %>%
  summarise(
    movies_rated = n(),
    avg_rating = mean(rating), 
    sd_rating = sd(rating, na.rm = TRUE)   
  )

print(user_summary)
```

# Boxplot of the movie ratings - use the raw dataset without the NaNs
```{r}
ggplot(rating_raw, aes(x = movie_title, y = rating, fill = movie_title)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 2.5, alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "Distribution of Movie Ratings",
    x = "Movie Title",
    y = "Rating"
  ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1) # Tilt
  )
```

## Findings and Recommendations
The low tech method of verbal survey resulted in a relatively small data set to analyze, with only 20 non-NaN rows. For the purposes of model development in the future, synthetic data generation may be utilized to expand the data set. 

For the ratings that were collected, The Odyssey had the highest average rating with a mean of 4.75. The movie with the lowest standard deviation in ratings was Toy Story 5 ($\sigma$ = 0.18) and highest standard deviation was Goat ($\sigma$ = 0.47) making them the least and most divisive movies respectively. 

Future work could include linking SQL database generation more directly with the R analysis, synthetic review data generation, and recommender model development.  


