---
title: "Project 2: Data Tidying"
subtitle: "Data Set N"
author: "Jocelyn Slater"
date:   today
format:
  html:
    toc: true
    code-fold: true
    theme: flatly
execute:
  echo:    true
  warning: false
editor: visual
---

```{r load-packages, message=FALSE, echo=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(tidyverse)
library(gt)
library(readr)
library(data.table)
library(stringr)
```

## Task (taken from Brightspace directly)

1. Objective
This project develops practical competency in transforming wide-format datasets into tidy formats suitable for downstream analysis. All transformations are performed using the tidyr and dplyr packages in R.

2. Dataset Selection
You must satisfy all three of the following requirements when selecting your datasets:
  - Select three (3) *distinct* wide-format datasets from the Week 5 Discussion 5A items.
  - You may substitute your own dataset only if it is structurally wide.
  - Do not use the instructor's Sample Post dataset.

3. Data Preparation
Complete all three sub-tasks below for each selected dataset. Steps 3.1 through 3.3 must be applied independently to every dataset.

3.1 Raw Data Construction
Create a source data file that captures all information from the original dataset while preserving its wide (untidy) structure. This file serves as the unmodified input to your transformation pipeline.

Create a .csv file containing all information from the original dataset.
Preserve the original wide format — do not tidy the data at this stage.
Alternative: store the dataset in a relational database instead of a flat .csv file.
 
DELIVERABLE Commit the raw dataset file(s) to your GitHub repository before proceeding to Step 3.2.

3.2 Data Import and Tidying
Primary Grading Component

Import your data into R and apply a reproducible sequence of tidyr and dplyr operations. Your code must demonstrate each of the following transformations explicitly:

- Reshape from wide to tidy (long) format
- Normalize variable structure
- Rename variables to follow a consistent naming convention
- Address missing or inconsistent values with documented handling decisions
 
The following example illustrates a minimal tidying pipeline using pivot_longer():

```{r, eval = FALSE}
# Import raw data
raw_data <- read.csv("data/dataset_wide.csv")

# Reshape to tidy format
tidy_data <- raw_data |>
  pivot_longer(
    cols      = -id,
    names_to  = "variable",
    values_to = "value"
  ) |>
  rename_with(tolower) |>
  drop_na()
```

3.3 Analysis
Perform the specific analysis requested in the original Discussion 5A item for each dataset. All analysis must operate on the tidy dataset produced in Step 3.2 — do not analyze the raw wide-format data directly.

Include the following outputs as appropriate for each analysis:
- Summary tables with clearly labeled columns and units
- Visualizations with axis labels, titles, and legends
- A brief narrative interpreting each result in context

4. Documentation and Reproducibility
All code and narrative must be written in a Quarto Markdown (.Qmd) file. Quarto extends the R Markdown ecosystem with improved cross-referencing, theming, and multi-format output — see the quarto.org documentation if you are new to the format.


Your .Qmd file must document each of the following in order:

- Data source — where the dataset originated and how it was obtained
- Data structure before tidying — column names, row count, and a representative sample
- Transformation steps — commented, executable code with explanations of each decision
- Analytical methods — a description of the statistical or summarization approach applied
- Conclusions — a substantive interpretation of results with reference to specific values
 

5. Submission Requirements
Submit the following two URLs for each of your three datasets. All six links must be functional at the time of grading.

URL to the .Qmd source file in your GitHub repository
URL to the rendered, published output on rpubs.com

IMPORTANT All datasets and code must be publicly accessible via GitHub. Private repositories will not be graded. Verify that all links resolve without authentication before submitting.

Submission Checklist
Before submitting, confirm each item below:

Three independent wide-format datasets selected from Discussion 5A
Raw data files committed to GitHub (Step 3.1)
Tidying pipeline documented and executable in .Qmd (Step 3.2)
Analysis uses only the tidy dataset (Step 3.3)
GitHub repository is public and all links resolve
Published RPubs output is accessible without login

## Outline
This outline will be replicated for each data set. Three different Code Base quarto files will be submitted, following the same format:

- Introduction (explain data source)
  - Anastasiia Gmyrina US Census population data 
  - Esra Dogan GDP per capita data
  - Jocelyn Slater fictional business data
  - Data is hosted here: https://github.com/jocslater-code/DATA607/tree/main/Project2
- Load in Data (.csv)
  - Show raw data format (column names, row count, and a representative sample)
- Tidy/Transformation data — commented, executable code with explanations of each decision
  - Reshape from wide to tidy (long) format
  - Normalize variable structure
  - Rename variables to follow a consistent naming convention
  - Address missing or inconsistent values with documented handling decisions
- Analytical methods — a description of the statistical or summarization approach applied
  - Draw from recommended analyses from the post
  - Summary tables with clearly labeled columns and units
  - Visualizations with axis labels, titles, and legends
  - A brief narrative interpreting each result in context
- Conclusions 
  — a substantive interpretation of results with reference to specific values
 

## Anticipated Challenges

-   Don't want to get mixed up with different data sets - need to keep them straight
- 

## Submission Requirements

-   Post all code and data (SQL, R, Quarto) in a GitHub repository

-   Submit the GitHub link

-   RPubs publication
