---
title: "Project 2: Data Tidying"
subtitle: "US Census Population Data"
author: "Jocelyn Slater"
date:   today
format:
  html:
    toc: true
    code-fold: true
    theme: flatly
execute:
  echo:    true
  warning: false
editor: visual
---

```{r load-packages, message=FALSE, echo=FALSE}
library(tidyverse)
library(janitor)
library(ggplot2)
library(scales)
library(gt)
```

## Introduction

Anastasiia Gmyrina suggested looking into the US Census population data, specifically, the 10 states that saw the largest positive and negative population changes from 2020-2022. The raw data was wide and not formatted for analysis, so before exploring the relationships, we first tidyed the data. 


## Import Data

Import data from github hosted .csv and glimpse what we are working with. 
```{r load-in-data}
url <- "https://raw.githubusercontent.com/jocslater-code/DATA607/refs/heads/main/Project2/census_top10_states_2022_untidy.csv"
df_raw <- read.csv(url)
head(df_raw)
```

## Tidy Data
The data needs to be formatted for analysis. I first clean up the column names using the Janitor library, covert from wide to long, extract clear years, and reorder the columns.
```{r tidy-data}
df_tidy <- df_raw %>%
  # Janitor to clean up the column names
  clean_names() %>%
  
  # Convert to long format
  pivot_longer(
    cols = starts_with(c("april", "july")),
    names_to = "estimate_period",
    values_to = "population"
  ) %>%
  
  # Clean up the period column text and extract clean numeric years
  mutate(
    # Format labels cleanly (e.g., "July 1 2021" or "April 1 2020 Estimates Base")
    estimate_period = str_replace_all(estimate_period, "_", " ") %>% str_to_title(),
    
    # Optional: extract explicit numeric year into a separate column
    year = as.numeric(str_extract(estimate_period, "\\d{4}"))
  ) %>%
  
  # Reorder columns logically
  select(rank, geographic_area, estimate_period, year, population)

head(df_tidy)
```

## Analysis

```{r pop-over-time-plot}
# Reorder data frame so it is in descending order by 2020 population
df_plot <- df_tidy %>%
  mutate(
    # Reorder factor levels descending by pop_2020
    pop_2020 = population[year == 2020][match(geographic_area, geographic_area[year == 2020])],
    geographic_area = fct_reorder(geographic_area, pop_2020, .desc = TRUE)
  )

# Make the plot
ggplot(df_plot, aes(x = year, y = population / 1e6, color = geographic_area, group = geographic_area)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_x_continuous(breaks = c(2020, 2021, 2022)) +
  labs(
    title = "U.S. State Populations (2020–2022)",
    x = "Year",
    y = "Population (Millions)",
    color = "State"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    legend.position = "right"
  )
```

To calculate the percent changes, first we make a new data frame with just the start and ending years. Then we can calculate absolute change, percent change, and add a new column for gained/lost population status.
```{r pop-over-time-table}
# Summary dataframe with percent changes
pop_change_summary <- df_tidy %>%
  filter(year %in% c(2020, 2022)) %>%
  pivot_wider(
    id_cols = c(rank, geographic_area),
    names_from = year,
    names_prefix = "pop_",
    values_from = population
  ) %>%
  mutate(
    abs_change = pop_2022 - pop_2020,
    pct_change = (pop_2022 - pop_2020) / pop_2020,
  ) %>%
  arrange(desc(pct_change))

# 2. Pass directly into gt
table_gt <- pop_change_summary %>%
  gt() %>%
  tab_header(
    title = md("**U.S. State Population Change Summary (2020–2022)**")) %>%
  tab_spanner(label = "Population", columns = c(pop_2020, pop_2022)) %>%
  tab_spanner(label = "Net Change", columns = c(abs_change, pct_change)) %>%
  cols_label(
    rank = "Rank",
    geographic_area = "State",
    pop_2020 = "2020",
    pop_2022 = "2022 Estimate",
    abs_change = "Absolute",
    pct_change = "Percentage"
  ) %>%
  fmt_number(columns = c(pop_2020, pop_2022, abs_change), decimals = 0) %>%
  fmt_percent(columns = pct_change, decimals = 2) %>%
  cols_align(align = "left", columns = geographic_area) %>%
  # Conditional text
  tab_style(
    style = cell_text(color = "#1b5e20", weight = "bold"),
    locations = cells_body(columns = pct_change, rows = pct_change > 0)
  ) %>%
  tab_style(
    style = cell_text(color = "#b71c1c", weight = "bold"),
    locations = cells_body(columns = pct_change, rows = pct_change < 0)
  ) %>%
  tab_options(
    heading.title.font.size = px(18),
    column_labels.font.weight = "bold",
    table.border.top.color = "transparent",
    table.border.bottom.color = "black"
  )

table_gt
```
    
## Discussion and Next Steps

- 

