---
title: "Project 2: Data Tidying"
subtitle: "World Bank GDP Per Capita"
author: "Jocelyn Slater"
date:   today
format:
  html:
    toc: true
    code-fold: true
    theme: flatly
execute:
  echo:    true
  warning: false
editor: visual
---

```{r load-packages, message=FALSE, echo=FALSE}
library(tidyverse)
library(janitor)
library(ggplot2)
library(scales)
library(gt)
```

## Introducion

Esra Dogan suggested looking at the GDP per capita data from the World Bank. The raw data was very messy with lots of missing data, so before any analysis could be done, we first needed to tidy the dataframe. Then we investigated which countries had the fastest growth in recent years.

## Import Data

Import data from github hosted .csv and glimpse what we are working with. 
```{r load-in-data}
url <- "https://raw.githubusercontent.com/jocslater-code/DATA607/refs/heads/main/Project2/WorldBankGDP/API_NY.GDP.PCAP.CD_DS2_en_csv_v2_440845.csv"
df_raw <- read.csv(url, skip = 4)
head(df_raw)
```

## Tidy Data
The data needs to be formatted for analysis. I first cleaned up the column names using the Janitor library, coverted from wide to long, extracted clear years, and droped rows that have all NA values. Then I picked just the geographic data (regions) that I am interested in exploring. 
```{r tidy-data}
df_tidy <- df_raw %>%
  # Janitor to clean up the column names
  clean_names() %>%
  
  # Pivot all year columns starting with 'x' into long format
  pivot_longer(
    cols = starts_with("x"),
    names_to = "year",
    values_to = "value"
  ) %>%
  
  # Extract years and cast as numeric
  mutate(
    # Remove the 'x' or 'x_' prefix and convert to numeric year
    year = as.numeric(str_remove(year, "^x_?"))
  ) %>%
  
  # Drop NAs
  filter(!is.na(value))


df_mini <- df_tidy %>% 
  filter(country_name %in% c("Middle East, North Africa, Afghanistan & Pakistan", "East Asia & Pacific", "European Union", "Latin America & Caribbean","North America" , "Africa Eastern and Southern", "Africa Western and Central"))

df_mini$country_name[df_mini$country_name == "Middle East, North Africa, Afghanistan & Pakistan"] <- "MENA"
df_mini$country_name[df_mini$country_name == "Africa Eastern and Southern"] <- "Southern and Eastern Africa"
df_mini$country_name[df_mini$country_name == "Africa Western and Central"] <- "Central and Western Africa"

head(df_mini)


```

## Analysis
We plotted all of the data to get a qualitative sense of overall trends and created summary tables to get a quantitative view of the GDP per capita data. GDP is in current (2026) US dollars.
```{r country-summary}
country_summary <- df_mini %>%
  group_by(country_name, country_code) %>%
  summarise(
    start_year = min(year, na.rm = TRUE),
    end_year   = max(year, na.rm = TRUE),
    min_gdp    = min(value, na.rm = TRUE),
    max_gdp    = max(value, na.rm = TRUE),
    mean_gdp   = mean(value, na.rm = TRUE),
    latest_gdp = value[which.max(year)],
    .groups = "drop"
  )

country_summary$start_year<- NULL
country_summary$end_year<- NULL

country_summary %>%
  gt() %>%
  tab_header(
    title = "GDP Per Capita Summary by Region",
  ) %>%
  fmt_currency(
    columns = c(min_gdp, max_gdp, mean_gdp, latest_gdp),
    currency = "USD",
    decimals = 2) %>%
  cols_label(
    country_name = "Country",
    country_code = "ISO Code",
    min_gdp      = "Min GDP",
    max_gdp      = "Max GDP",
    mean_gdp     = "Mean GDP",
    latest_gdp   = "Latest GDP"
  )

```

```{r plot_gdp}
ggplot(df_mini, aes(x = year, y = value, color = country_name)) +
  geom_line(linewidth = 1) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "GDP Per Capita Over Time by Region",
    x = "Year",
    y = "GDP per Capita (Current US$)",
    color = "Region"
  ) +
  theme_minimal()
```
I also want to calculate the percent change in GDP to see which region has grown the most over the time period
```{r}
pct_change_df <- df_mini %>%
  arrange(country_name, year) %>%
  group_by(country_name) %>%
  summarize(
    pct_change = ((last(value) - first(value)) / first(value)) * 100
  ) %>%
  arrange(desc(pct_change))

# Great table
pct_change_df %>%
  gt() %>%
  tab_header(
    title = "GDP Percent Change by Region",
  ) %>%
  fmt_number(
    columns = pct_change,
    decimals = 0
  ) %>%
  cols_label(
    country_name = "Country",
    pct_change = "Percent Change in GDP (%)"
  )

```
## Discussion and Next Steps

The region with the largest percent change in GDP was East Asia & Pacific (9,173%) while the region with the smallest is Southern and Eastern Africa. 

The next steps could be to expand this analysis to new regions, look at all countries, or break East Asia & Pacific down by individual countries.
