---
title: "Project 2: Data Tidying"
subtitle: "Fictional Business Data"
author: "Jocelyn Slater"
date:   today
format:
  html:
    toc: true
    code-fold: true
    theme: flatly
execute:
  echo:    true
  warning: false
editor: visual
---

```{r load-packages, message=FALSE, echo=FALSE}
library(tidyverse)
library(janitor)
library(ggplot2)
library(scales)
library(dplyr)
library(tidyr)
```

## Introducation

I created a fictional business data .csv file containing very messy data. The raw data was wide with inconsistent units and an outlier. It was not formatted for analysis, so before exploring the relationships, I first tidyed the data. I then calculated profit and found the city with the greatest earnings. 

## Import Data

Import data from github hosted .csv and glimpse what we are working with. 
```{r load-in-data}
url <- "https://raw.githubusercontent.com/jocslater-code/DATA607/refs/heads/main/Project2/5A_BusinessCostsRevenue.csv"
df_raw <- read.csv(url)
glimpse(df_raw)
```

## Tidy Data
The data needs to be formatted for analysis. I first clean up the column names using the Janitor library. In order to pivot and combine columns, I need to convert everything to characters and remove the commas. There are also some months that are reported in thousands, so the units of the cost and revenue need to be standardized. Then I can covert from wide to long and fix the column names
```{r tidy-data}
df_tidy <- df_raw %>%

  clean_names() %>%
  
  # Convert everything to character to prevent mismatches later
  mutate(across(everything(), as.character)) %>%
  
  # Pivot all month columns 
  pivot_longer(
    cols = -c(location, name),
    names_to = "measure",
    values_to = "value"
  ) %>%
  
  # Remove commas and convert to to numeric
  mutate(
    value = as.numeric(str_remove_all(value, ","))
  ) %>%
  
  # Extract month name and cost/revenue
  mutate(
    month = str_extract(measure, "^[a-z]+"),
    metric = if_else(str_detect(measure, "cost"), "cost", "revenue"),
    
    # Scale cost values if given in thousands
    value = if_else(metric == "cost" & value <= 100, value * 1000, value)) %>%


  # Pivot into separate `cost` and `revenue` columns
  pivot_wider(
    id_cols = c(location, name, month),
    names_from = metric,
    values_from = value
  ) %>%
  
  # Clean up month formatting and reorder
  mutate(
    month = str_to_title(month),
    month = factor(month, levels = c("Jan", "Feb", "March", "April", "May", "June"))
  ) %>%
  arrange(location, month)

head(df_tidy)
```

## Analysis

First we subtract the costs from revenue to calculate profit and summarize that in a table. This allows us to see that there is an outlier in the data. The Boston revenue for May was \$6,150,000, suggesting a typo. When that is corrected to \$61,500, the data looks much more reasonable. 
```{r calcualte-profit}
df_tidy <- df_tidy %>%
  mutate(
    profit = revenue - cost,
    month = factor(month, levels = c("Jan", "Feb", "March", "April", "May", "June"))
  )

# Calculate summary by location
df_summary <- df_tidy %>%
  group_by(location) %>%
  summarise(
    Total_Cost = sum(cost, na.rm = TRUE),
    Total_Revenue = sum(revenue, na.rm = TRUE),
    Total_Profit = sum(profit, na.rm = TRUE),
    Avg_Profit_Margin = mean(profit / revenue, na.rm = TRUE)
  )

df_summary
```

```{r fix-outlier}
# Fix the typo
df_tidy <- df_tidy %>%
  mutate(revenue = if_else(location == "Boston" & month == "May", 61500, revenue))

# Re-calculate profit
df_tidy <- df_tidy %>%
  mutate(
    profit = revenue - cost,
    month = factor(month, levels = c("Jan", "Feb", "March", "April", "May", "June"))
  )

# Re-calculate summary by location
df_summary <- df_tidy %>%
  group_by(location) %>%
  summarise(
    Total_Cost = sum(cost, na.rm = TRUE),
    Total_Revenue = sum(revenue, na.rm = TRUE),
    Total_Profit = sum(profit, na.rm = TRUE),
    Avg_Profit_Margin = mean(profit / revenue, na.rm = TRUE)
  )

# Build Great Table
df_summary %>%
  gt() %>%
  tab_header(
    title = md("**Financial Performance by Location**"),
    subtitle = md("*Summary of Total Cost, Revenue, and Profit*")
  ) %>%
  cols_label(
    location = "Location",
    Total_Cost = "Total Cost",
    Total_Revenue = "Total Revenue",
    Total_Profit = "Total Profit",
    Avg_Profit_Margin = "Avg Profit"
  ) %>%
  fmt_currency(
    columns = c(Total_Cost, Total_Revenue, Total_Profit),
    currency = "USD",
    decimals = 0
  ) %>%
  fmt_percent(
    columns = Avg_Profit_Margin,
    decimals = 1
  ) %>%
  grand_summary_rows(
    columns = c(Total_Cost, Total_Revenue, Total_Profit),
    fns = list(Total = ~sum(.)),
    fmt = ~ fmt_currency(., currency = "USD", decimals = 0)
  ) 



```


```{r plot}
ggplot(df_tidy, aes(x = month, y = profit, color = location, group = location)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = scales::dollar) +
  theme_minimal() +
  labs(
    title = "Monthly Profit by Location",
    x = "Month",
    y = "Profit",
    color = "Location"
  )

```

## Discussion and Next Steps
After the Boston outlier was identified and corrected, Boston still remained the second best performer. However, Dallas just out performed Boston with a 24% profit margin compared to Boston's 21%. Both the cities owe their status to one outstanding month, rather than steady high profits. Chicago shows volitile month to month performance, while the rest of the cities earned more modestly, but also more consistently. 

Next steps with this data could be to aggregating all costs/revenues to see the national performance over the 6 months.


